\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{xcolor}
\title{ProbOS --- Month 1 Retrospective}
\author{Nisong Monyimba}
\date{\today}

\begin{document}
\maketitle

\section*{Period}
Week 1 through Week 4. Status at close: kernel foundation complete, PDSL
v0.1 shipped, 3 cross-discipline examples validated, CI fully green.

\section*{Final numbers}
\begin{itemize}[leftmargin=*]
  \item Python tests: 264/264 passing
  \item C++ tests: 13/13 passing
  \item mypy strict: 0 errors
  \item ruff: 0 warnings
  \item CI jobs: 4/4 green (Python, C++, LaTeX manuscript, full
    integration)
\end{itemize}

\section*{What shipped}

\subsection*{Week 1 --- Distribution ABC}
\begin{itemize}[leftmargin=*]
  \item \texttt{Distribution} ABC with \texttt{sample()}, \texttt{pdf()},
    \texttt{log\_pdf()}, \texttt{cdf()}, \texttt{ppf()}
  \item Concrete distributions: \texttt{Normal}, \texttt{LogNormal},
    \texttt{Beta}, \texttt{Uniform}
  \item 44 Python tests, 13 C++ tests
\end{itemize}

\subsection*{Week 2 --- Model ABC + Physics}
\begin{itemize}[leftmargin=*]
  \item \texttt{Model} ABC defining the \texttt{forward\_batch()} contract
  \item \texttt{BatteryModel2Cell}: 8-state Arrhenius thermal-runaway
    ODE, validated against Kim et al. (2007)
  \item CLT convergence demo, deterministic ODE demo
\end{itemize}

\subsection*{Week 3 --- Monte Carlo Kernel}
\begin{itemize}[leftmargin=*]
  \item \texttt{MonteCarloEngine}: vectorised particle propagation,
    percentile tracking
  \item \texttt{SobolSensitivity}: Sobol S1/ST indices via SALib
  \item \texttt{ProvenanceTracker}: causal audit trail for individual
    particle trajectories
  \item 91 new tests across MC engine, sensitivity, provenance
\end{itemize}

\subsection*{Week 4 --- Speed, Language, Cross-Discipline}
\begin{itemize}[leftmargin=*]
  \item \textbf{Monday:} DP profiling. Real wins: 1040x SALib cache, 45x
    CLT subset trick. One real non-win (inv\_RT precompute $\sim$1x)
    documented honestly.
  \item \textbf{Tuesday:} C++ \texttt{BatteryCell} + OpenMP
    \texttt{MonteCarloEngineOMP}. Confirmed 7x speedup over Python
    serial via genuine benchmark.
  \item \textbf{Wednesday:} PDSL compiler v0.1. Lark grammar to AST to
    codegen to compiler, 28 tests.
  \item \textbf{Literature pass:} added Feng 2018, Saltelli 2019, Chopin
    and Papaspiliopoulos 2020, Naesseth 2019, FDA 2019.
  \item \textbf{Repo hygiene:} moved generated PNGs into
    \texttt{outputs/figures/}, fixed a CI break (missing \texttt{lark}
    dependency), wrote \texttt{check\_ci.sh}.
  \item \textbf{Saturday:} 3 cross-discipline examples proving kernel
    domain-agnosticism (option pricer, ED queue, clinical trial).
  \item \texttt{docs/study/study\_guide.md} created as a permanent
    running study guide.
\end{itemize}

\section*{What went right}
\begin{enumerate}[leftmargin=*]
  \item Honest profiling over claimed wins. The DP session found
    inv\_RT precomputation gave no real speedup and said so directly.
  \item Kernel domain-agnosticism validated, not assumed. Three Week 4
    examples needed zero kernel changes, only new \texttt{Model}
    subclasses.
  \item CI treated as a first-class deliverable. The \texttt{lark}
    dependency break was caught and fixed same-session.
  \item Test discipline held under example-code growth. Even demo
    scripts got real test suites (34 tests total).
\end{enumerate}

\section*{What was harder than expected}
\begin{enumerate}[leftmargin=*]
  \item PDSL parser debugging. Unary minus inside function calls
    required several grammar iterations.
  \item Repo hygiene delayed feature work. PNG reorganisation and the
    resulting CI break took real Week 4 time.
  \item Terminal and heredoc fragility. Shell history expansion and
    long heredocs breaking mid-stream cost debugging cycles.
\end{enumerate}

\section*{Carry-forward risks for Month 2}
\begin{itemize}[leftmargin=*]
  \item No FastAPI or pybind11 yet. The kernel is pure-Python plus a
    separate C++ binary.
  \item Particle filter is unbuilt. \texttt{ProvenanceTracker} gives
    causal post-hoc analysis but no online inference yet.
  \item PDSL v0.1 has no control flow. Current grammar supports linear
    drift/diffusion declarations only.
\end{itemize}

\end{document}
