\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{xcolor}
\title{ProbOS --- Month 3, Week 13\\PDSL v0.2: Control Flow}
\author{Nisong Monyimba}
\date{\today}

\begin{document}
\maketitle

\section*{Week 13 goal}
Extend PDSL from v0.1 to v0.2 by adding comparison operators and
conditional expressions -- the first control-flow constructs PDSL
has ever supported.

\textbf{Scope confirmed via direct inspection of \texttt{grammar.lark}
before writing this plan:} PDSL v0.1 has ZERO comparison operators
and ZERO conditional constructs -- only arithmetic (+, -, *, /, **,
unary +/-) and function calls.

\textbf{Design decision, stated up front:} this adds
EXPRESSION-level conditionals (\texttt{if cond then expr else expr}),
NOT statement-level control flow or loops. PDSL describes a model's
per-step drift declaratively; the outer simulation engine already
handles the step-by-step loop. An expression-level conditional (for
clamping, thresholding, or piecewise physical behavior) is the right
granularity.

\section*{Day-by-day plan}

\subsection*{Day 1 -- Grammar design}
\begin{itemize}[leftmargin=*]
  \item Add comparison operators to \texttt{grammar.lark}: \texttt{<},
    \texttt{>}, \texttt{<=}, \texttt{>=}, \texttt{==}, \texttt{!=} --
    at a new precedence level between \texttt{expr} and a new
    top-level \texttt{condition} rule
  \item Add a conditional expression rule:
    \texttt{"if" condition "then" expr "else" expr -> conditional}
  \item Test the grammar in isolation (Lark parse-tree inspection)
    before touching \texttt{parser.py} or \texttt{codegen.py}
\end{itemize}

\subsection*{Day 2 -- AST nodes and parser}
\begin{itemize}[leftmargin=*]
  \item Add \texttt{Comparison} and \texttt{Conditional} AST node
    types to \texttt{ast\_nodes.py}, matching existing node style
  \item Extend \texttt{parser.py} to build these new AST nodes
\end{itemize}

\subsection*{Day 3 -- Code generation}
\begin{itemize}[leftmargin=*]
  \item Extend \texttt{codegen.py} to emit correct, VECTORISED
    Python: a naive \texttt{if/else} would not vectorise correctly
    over N particles -- must generate \texttt{xp.where(cond,
    expr\_true, expr\_false)}, matching the vectorisation discipline
    already used throughout the hand-written kernel models
\end{itemize}

\subsection*{Day 4 -- Test coverage}
\begin{itemize}[leftmargin=*]
  \item Grammar-level, codegen-level, and end-to-end tests: a
    genuinely new PDSL model using a conditional compiles and runs
    correctly through the full pipeline, matching a hand-written
    Python equivalent
\end{itemize}

\subsection*{Day 5 -- Existing model regression check}
\begin{itemize}[leftmargin=*]
  \item Confirm ALL existing PDSL v0.1 models/examples still parse
    and compile correctly -- strictly additive, not breaking
\end{itemize}

\subsection*{Day 6 -- pre\_week\_audit.sh + CI}
\begin{itemize}[leftmargin=*]
  \item Full audit pass
\end{itemize}

\subsection*{Day 7 -- Documentation + retrospective}
\begin{itemize}[leftmargin=*]
  \item Update \texttt{docs/vision/main.tex} if it references PDSL
    version status
  \item Retrospective appended once complete
\end{itemize}

\section*{Exit criteria}
PDSL v0.2 correctly parses, compiles, and executes conditional
expressions with comparison operators, generating correctly-
vectorised NumPy/CuPy code (\texttt{xp.where}), with zero regressions
to any existing v0.1 model.


\section*{What was actually accomplished (retrospective)}

\begin{itemize}[leftmargin=*]
  \item PDSL v0.2 shipped: comparison operators (<, >, <=, >=, ==,
    !=) and expression-level conditionals (if cond then expr else
    expr) -- the first control-flow constructs PDSL has ever
    supported.
  \item An honest correction was made to this plan document's own
    stated intention during Day 3: the plan said codegen ``must
    generate xp.where(), matching CuPy dispatch'' -- but direct
    inspection of the actual codegen.py showed PDSL v0.1 has no
    xp/CuPy dispatch at all, only plain \texttt{np.}. np.where() was
    the correct, properly-scoped choice.
  \item Genuine end-to-end proof of correct vectorisation: 5
    particles above a threshold and 5 below, in the SAME
    \texttt{forward\_batch()} call, correctly diverged based on their
    own individual state values. A boundary test confirmed strict
    comparison semantics.
  \item Two real, unrelated bugs were found and fixed, surfaced by
    \texttt{check\_ci.sh}'s corrected, accurate failure reporting:
  \begin{enumerate}[leftmargin=*]
    \item \texttt{check\_ci.sh} itself had a genuine race condition,
      permanently fixed by matching the exact commit SHA with a
      polling retry loop.
    \item CI's C++ dependency installation failed due to a broken
      Microsoft azure-cli apt repository -- a genuine, external
      infrastructure issue. Fixed robustly (grep by repo file
      CONTENT, not a guessed filename).
  \end{enumerate}
\end{itemize}

\section*{What went right}
\begin{enumerate}[leftmargin=*]
  \item Verifying assumptions against the actual current codebase
    caught the xp/CuPy scope-creep risk before any code was written.
  \item A test failure was correctly diagnosed as being in the TEST
    SCRIPT's PDSL syntax, not the actual codegen fix.
  \item check\_ci.sh's fix immediately proved its own value: it
    correctly caught a genuine CI failure, leading directly to
    finding and fixing the Microsoft-repo infrastructure issue.
\end{enumerate}

\section*{Numbers at Week 13 close}
\begin{itemize}[leftmargin=*]
  \item Python tests: 408/408 passing
  \item mypy strict: 0 errors
  \item ruff: 0 warnings
  \item Zero regressions to any existing PDSL v0.1 model
\end{itemize}
\end{document}
