\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{xcolor}
\title{ProbOS --- Month 3 Plan}
\author{Nisong Monyimba}
\date{\today}

\begin{document}
\maketitle

\section*{Status}
Per \texttt{docs/monthly\_plans/overall/main.tex}, Month 3 was always
DIRECTIONAL rather than pre-specified in detail -- Month 1's
retrospective explicitly flagged over-specifying future months as a
mistake, corrected by deferring detailed week-by-week planning until
the month is actually about to begin. This document intentionally
states THEMES, not a committed day-by-day schedule; the latter will
be written (following the same plan-then-retrospective process used
for every week so far) once Month 3's actual work begins.

\section*{Candidate themes, in rough priority order}

\subsection*{GPU kernel path}
Referenced as a genuine open item in
\texttt{docs/sbir\_phase1\_onepager.md}'s SBIR Ask (``GPU kernel, 10x
speedup over CPU, building on the existing C++/OpenMP path''). The
Month 1 Week 4 C++/OpenMP kernel achieved a measured 7x speedup over
the Python engine; a GPU path would need its own honest profiling
discipline (the same ``measure, don't assume'' standard that found
DP's real 1040x/45x wins alongside one genuine non-win in Month 1
Week 4) before any speedup number is claimed.

\subsection*{Extend the REST API's model registry -- DONE (Week 9)}
Completed: \texttt{python/server/main.py} now registers all four
models (battery, option pricer, ED queue, clinical trial), with
\texttt{/filter} correctly and honestly rejecting \texttt{clinical\_trial}
rather than returning a meaningless result. See
\texttt{docs/monthly\_plans/month3/week9/main.tex} for the full
retrospective.

\subsection*{Pharmaceutical stability model -- NEW, near-term priority}
Per the strategic reassessment documented in
\texttt{docs/vision/main.tex}: drug degradation kinetics are
Arrhenius reaction kinetics -- the exact same mathematical
structure as \texttt{BatteryModel2Cell}'s SEI decomposition, with
drug potency as the depleting quantity in place of battery
capacity. An estimated 80\% of the existing model architecture
transfers directly; remaining work is domain-appropriate parameter
priors and validation against a real stability-study dataset,
matching the same discipline used for Kim (2007). This directly
serves FDA CMC / ICH Q1A requirements, the same regulatory pattern
already validated for battery safety.

\subsection*{Automotive / EV safety positioning -- NEW, near-term priority}
Not new engineering -- the existing, already-validated battery
model, reframed for ISO 26262's hazard-analysis and documented-
failure-mode requirements, which are structurally analogous to
what FDA/NRC already require. The primary remaining work is
outreach and positioning specific to automotive/EV safety
engineers, not new model development.

\subsection*{PDSL v0.2 -- control flow}
Month 1 Week 4's PDSL compiler v0.1 supports only flat drift/diffusion
declarations, no loops or conditionals. Genuinely useful, but not
urgent -- no current model has needed it yet.

\subsection*{Apply the particle filter to real (non-synthetic) data}
The most consequential open SBIR ask (``apply the particle filter to
real sensor data from a pilot partner''), and the one most dependent
on external factors (an actual pilot partnership) rather than purely
technical work. Cannot be scheduled with certainty; flagged here so it
is not forgotten.

\subsection*{Continued quality-standard maintenance}
Per the mid-Month-2 hardening pass: \texttt{pre\_week\_audit.sh} and
\texttt{docs/standards/quality\_standards.md} are living documents.
Any new gap discovered (declared-but-unenforced checks, stale
documentation, vestigial code patterns) should be closed in the same
session it is found, not deferred -- the standing practice established
this month, to be continued.

\section*{Process}
As with every prior month, Month 3's actual week-by-week plan will be
written and compiled to PDF (\texttt{docs/monthly\_plans/month3/weekN/main.tex})
before that week's work begins, with a retrospective appended once it
completes. This document will be revisited and made more concrete once
Month 3 actually starts.

\end{document}
